---
title: "Monumentality by Numbers"
author:
  - Hageneuer, Sebastian
  - Schmidt, Sophie
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
keywords: |
  quantification; regression analysis; size and effort; monumentality
---


```{r include=FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  include = TRUE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  fig.width = 8, 
  fig.height = 6
)
```

## intro

This R package offers the code to the calculations described in:
Hageneuer, S. / Schmidt, S., Monumentality by Numbers, in: Hageneuer et al. (Hrsg.), Size Matters - Understanding Monumentality across ancient Civilizations (transcript-Verlag Bielefeld 2019).

```{r load data}
uebersicht <- read.csv2("../data/raw_data/uebersicht3.csv", sep = "\t", dec=",",  fileEncoding="UTF-8")

# load needed packages
library(tidyverse)
library(ggrepel)
library(ggpmisc)
library(RColorBrewer)
library(reshape2)
library(scales)
library(mgcv)
```

# Effort calculation

Aim of this part: Sum the different materials (e.g. differently coloured ceramic cones) that are going to be analysed together and calculate the effort, which is dependant on the volume of different materials. Sum volume and effort of each building.

```{r effort}

# summing data, create table with all relevant volumes of material categories

aufwand <- data.frame(uebersicht$X, uebersicht$Gebaeude, uebersicht$Mriemchen + uebersicht$Mlehmziegel, uebersicht$Mlehm, uebersicht$Mlehm_g + uebersicht$Mlehm_w,  uebersicht$Mschilf, uebersicht$Mholz, uebersicht$Masphalt + uebersicht$Kalksteinplatten, (uebersicht$Mkeramikstifte_r + uebersicht$Mkeramikstifte_schw + uebersicht$Mkeramikstifte_w + uebersicht$Mkeramikstifte_g + uebersicht$Mflaschen), uebersicht$MasseSteinstifte, uebersicht$Mkunststein)
colnames(aufwand)[3] <- "Mziegel" # Riemchen sind Ziegel!   # rename column to Mziegel = mass of mud bricks
colnames(aufwand)[5] <- "MFarblehm" # gelber und weißer     # rename column to MFarblehm = mass of coloured clay
colnames(aufwand)[8] <- "MsonstMat" # Apshalt und Kalkstein    # rename column to MsonstMat = mass of other materials
colnames(aufwand)[9] <- "MKeramikZier" # Keramische Zierelemente: Flaschen und Keramikstifte # rename column to MKeramikzier = mass of ceramic decoration
colnames(aufwand)[10] <- "MSteinstifte" # Steinstifte untersch. Farben    # rename column to MSteinstifte = mass of stone cones

# sum the volumes
aufwand$summe_volumen <- aufwand$Mziegel + aufwand$uebersicht.Mlehm + aufwand$uebersicht.Mschilf + aufwand$uebersicht.Mholz + aufwand$MsonstMat + aufwand$MSteinstifte + aufwand$MKeramikZier + aufwand$MFarblehm + uebersicht$Mkunststein

# multiply with effort factor
aufwand$Lehm <- aufwand$uebersicht.Mlehm*1
aufwand$Schilf <- aufwand$uebersicht.Mschilf*1
aufwand$Farblehm <- aufwand$MFarblehm*2
aufwand$Ziegel <- aufwand$Mziegel*4
aufwand$KeramikZier <- aufwand$MKeramikZier*4
aufwand$holz <- aufwand$uebersicht.Mholz*7
aufwand$Steinstifte <- aufwand$MSteinstifte*8
aufwand$sonstMat <- aufwand$MsonstMat*12
aufwand$Kunststein<- uebersicht$Mkunststein*16

# sum all effort
aufwand$summe_aufwand <- aufwand$Lehm + aufwand$Schilf + aufwand$Farblehm  + aufwand$Ziegel + aufwand$holz + aufwand$KeramikZier  +  aufwand$sonstMat + aufwand$Steinstifte + aufwand$Kunststein


```



# Analyse the middle-hall buildings quantitatively
Aim of the following code chunk: We subsetted a larger data frame for just the tripartite buildings. Calculations later showed we need the data frame of tripartite buildings without the stone cone building as well.



```{r middle-hall-buildings}
## subset and remove Stone-Cone Building

aufwand_MSG <- subset(aufwand, uebersicht.Gebaeude == "Mittelsaalgebaeude") 
aufwand_MSG_oSSG <- aufwand_MSG[-9,] ## remove stone-cone building 

```


## amount of material
Aim: Create a stacked bar plot, in which we can see the amount of different materials used for each building. Absolute numbers used so the stack shows the absolute size of the building.

```{r compare buildings for amount of materials used}

# remove unwanted data: effort calculations not needed

vol_MSG3 <- aufwand_MSG
vol_MSG3$category <- row.names(vol_MSG3)
vol_MSG3 <- vol_MSG3[,1:12]

# melt dataframe

vol_MSG3_m <- melt(vol_MSG3, id.vars = c("uebersicht.X", "uebersicht.Gebaeude", "summe_volumen"))


vol_MSG3_m <- vol_MSG3_m %>%
  arrange(summe_volumen, uebersicht.X) %>%               # sort the dataframe for its size
  mutate(uebersicht.X = factor(uebersicht.X, unique(uebersicht.X))) # change factor levels to how they've been arranged above

vol_MSG3_m$variable <- ordered(vol_MSG3_m$variable, levels = c("Mziegel", "uebersicht.Mlehm", "MFarblehm", "uebersicht.Mschilf",  "uebersicht.Mholz", "MKeramikZier", "MSteinstifte", "MsonstMat", "uebersicht.Mkunststein")) # order variables for the stacked bar plot so they are always in the same order (to compare with the following bar plot)


ggplot(data = vol_MSG3_m)+
  geom_bar(aes(x = uebersicht.X, y = value, fill = variable),
           stat = "identity",
           position = "stack")+
   scale_y_continuous("Size of the building (cubic meter)")+
  labs(x = "")+
      theme_bw()+
    theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 12))+
    scale_fill_brewer(palette = "Set1",
                    breaks = c("Mziegel", "uebersicht.Mlehm", "MFarblehm", "uebersicht.Mschilf",  "uebersicht.Mholz", "MKeramikZier", "MSteinstifte", "MsonstMat", "uebersicht.Mkunststein"),
                    labels = c("Mud bricks", "Clay",  "Colored Clay", "Reed", "Timber", "Ceramic Decoration", "Others", "Stone Cones", "Cast Stone"))+
  guides(fill=guide_legend(title=NULL))

ggsave("../figures/Mittelsaalgebaeude_MASSE.eps", width = 20, height = 15, units = "cm") # save plot for publication

```


## How much effort do the materials imply?
Aim: Create a stacked bar plot, in which we can see the effort of the different materials used for each building. Percentage used so the stack shows composition of material effort without regard to absolute size of the building.

```{r Compare different amounts of effort regarding material}

aufwand_MSG2 <- aufwand_MSG
#remove unwanted info: no volumes needed
aufwand_MSG2$category <- row.names(aufwand_MSG2)
aufwand_MSG2 <- select(aufwand_MSG2, -starts_with ("uebersicht.M"))
aufwand_MSG2 <- select(aufwand_MSG2, -starts_with ("M"))

aufwand_MSG2 <- aufwand_MSG2[,1:12]

# melt data 

aufwand_MSG_m <- melt(aufwand_MSG2, id.vars = c("uebersicht.X", "uebersicht.Gebaeude", "summe_volumen"))

aufwand_MSG_m <- aufwand_MSG_m %>%
  arrange(summe_volumen, uebersicht.X, variable) %>%               # sort the dataframe for its size
  mutate(uebersicht.X = factor(uebersicht.X, unique(uebersicht.X))) # change factor levels to how they've been arranged above

aufwand_MSG_m$variable <- ordered(aufwand_MSG_m$variable, levels = c("Ziegel","Lehm", "Farblehm", "Schilf", "holz", "KeramikZier", "sonstMat", "Steinstifte", "Kunststein")) # order variables for the stacked bar plot so they are always in the same order (to compare with the bar plot before)


ggplot(data = aufwand_MSG_m)+
  geom_bar(aes(x = uebersicht.X, y = value, fill = variable),
           stat = "identity",
           position = "fill")+
   scale_y_continuous("Relative effort", labels = percent_format())+
  theme_bw()+
  scale_fill_brewer(palette = "Set1",
                    breaks =c("Ziegel","Lehm", "Farblehm", "Schilf", "holz", "KeramikZier", "sonstMat", "Steinstifte", "Kunststein"),
                    labels = c("Mud bricks", "Clay",  "Colored Clay", "Reed", "Timber", "Ceramic Decoration", "Others", "Stone Cones", "Cast Stone"))+
  labs(x = "")+
    theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 12))+
  guides(fill=guide_legend(title=NULL))

ggsave("../figures/Mittelsaalgebaeude_AUFWAND.eps", width = 20, height = 15, units = "cm") #save for publication


```


# Regression analysis

Aim: Analyse the relationship between size and effort.
Starting hypothesis: linear relatoionship between size and effort. This gives a fairly good R²-value, but could be better. The stone-cone building doesn't fit well. Also the residual analysis (see below) shows there's a relationship between residuals and size. Therefore two options:
1. linear relationship without stone cone building
2. test non-linear regressions

## Linear regressions

```{r Diagram linear regression, fig.height=10, fig.width=12}

ggplot(data = aufwand_MSG)+
  geom_point(aes(x = summe_volumen,            # point cloud all middle-hall-buildings
                 y = summe_aufwand))+
  geom_smooth(data = aufwand_MSG,           # regression all middle-hall-buildings -> bad fit
              method='lm',
              se = FALSE,      # no confidence intervall
              aes(x = summe_volumen,
                  y = summe_aufwand, 
                  colour = "black" )) + 
  stat_poly_eq(data = aufwand_MSG,          # Information on regression all middle hall buidlings
               aes(x = summe_volumen,
                   y = summe_aufwand,
                   label =  paste(..eq.label..,  # equation of regression line
                                  ..adj.rr.label..,   # adjusted R-squared
                                  sep = "~~~~")),
               rr.digits = 4,  # 4 digits behind point
               formula = y~x, 
               parse = TRUE,
               size = 5,
               color = "black",                  
               label.y.npc = 0.8) +
       stat_fit_glance(data = aufwand_MSG, # p value of regression all middle hall buildings
                aes(x = summe_volumen,
                    y = summe_aufwand, 
                    label = paste("p-value: ",signif(..p.value.., digits = 4))),
                    label.y.npc = 0.8,
                  size = 5,
                   color = "black", 
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text" ) +
   geom_smooth(data = aufwand_MSG_oSSG,
              method='lm',                      # regression for middle-hall-buildings withouth stone cone building
              se = FALSE,      ## no confidence intervall
              aes(x = summe_volumen,
                   y = summe_aufwand, colour = "gray50")) +
  stat_poly_eq(data = aufwand_MSG_oSSG,           # info on regression for middle-hall buildings without stone-cone building
               aes(x = summe_volumen,
                    y = summe_aufwand,
                   label =  paste(..eq.label.., 
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               rr.digits = 4,
               formula = y~x, 
               parse = TRUE,
               size = 5,
               color = "gray50",
               label.y.npc = 0.7)+            # Y-coordinate for this info
     stat_fit_glance(data = aufwand_MSG_oSSG,  # p-value
                aes(x = summe_volumen,
                    y = summe_aufwand, 
                    label = paste("p-value: ",signif(..p.value.., digits = 5))),
                    label.y.npc = 0.7,
                  size = 5,
                   color = "gray50", 
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text")+               
  geom_label_repel(aes(x = summe_volumen,  # label the points
                        y = summe_aufwand,
                       label = uebersicht.X),
                   box.padding = 0.35,             # bounding box of label
                   point.padding = 0.5,
                   size = 5,
                   nudge_x = 0,
                   nudge_y = 0.05,                 
                   direction = "y")+
  scale_colour_manual(name="Linear model",
                      values=c("black", "gray50"),
                      breaks=c("black", "gray50"),
                      labels=c("for all middle-hall buildings", "without the Stone-Cone Building"))+
  labs(y = "Relative effort index", 
       x = expression(Size~of~buidings~(m^{3})))+
  theme(text = element_text(size=15))+
   theme(legend.position="bottom", legend.text = element_text(size = 12))

ggsave("../figures/Mittelsaalgebäude_vgl.eps", width = 20, height = 20, units = "cm")


```

 

## Non-linear regressions
Aim: Test different non-linear regressions in comparison to linear regression model (of all tripartite buildings). Two chosen: LOESS because it supposedly doesn't weigh outliers so much (Baxter 2015: 80-82) and GAM, with a relative simple y ~ poly(x, 2)-formula. Visualisation and summary of both methods for evaluation.

```{r non-linear regression comparison}
# comparison of different regressions of all middle hall buildings


ggplot(data = aufwand_MSG)+
  geom_point(aes(x = summe_volumen,            # point cloud
                 y = summe_aufwand))+
  geom_smooth(data = aufwand_MSG,           # linear regression for middle-hall buildings
              method='lm',
              se = FALSE,     
              aes(x = summe_volumen,
                  y = summe_aufwand,
                  colour = "black" )) +
    geom_smooth(data = aufwand_MSG,
              method='loess',                      # not-linear regression for all middle hall buildins: loess
              se = TRUE,      # confidence intervall 
              aes(x = summe_volumen,
                  y = summe_aufwand, colour = "gray50")) +
  geom_smooth(data = aufwand_MSG,
              method='gam',                      # not-linear regression for all middle hall buildins: gam
              formula = y ~ poly(x, 2),
              se = TRUE,      # confidence intervall yes please
              aes(x = summe_volumen,
                  y = summe_aufwand, colour = "gray20")) +
    geom_label_repel(aes(x = summe_volumen,  
                       y = summe_aufwand,
                       label = uebersicht.X),
                   box.padding = 0.35,             
                   point.padding = 0.5,
                   size = 5,
                   nudge_x = 0,
                   nudge_y = 0.05,                 
                   direction = "y")+
  scale_colour_manual(name="regression model",
                      values=c("black", "gray50", "gray20"),
                      breaks=c("black", "gray50", "gray20"),
                      labels=c("linear model", "loess", "gam"))+
  labs(y = "Relative effort index",
       x = expression(Size~of~buidings~(m^{3})))+
  theme(text = element_text(size=15))+
  theme(legend.position="bottom", legend.text = element_text(size = 12))

fit3 <- loess(summe_aufwand ~ summe_volumen, data = aufwand_MSG)

fit4 <- gam(formula = summe_aufwand  ~ poly(summe_volumen, 2), data = aufwand_MSG)

summary(fit3)

summary(fit4)
```


LOESS: very large confidence intervall needed and the form of the regression does not seem to be logical (why should the effort fall again at the end?)
GAM: R²-value of 0.72 even worse than the one of the linear regression.
conclusion: both not very helpful.


## Residual analysis
Aim: Look at the residuals of both linear regressions to evaluate them better. 

1. for all tripartite buildings:

First plot visualises the residuals on the original plot. 

Residuals vs Fitted plot shows theres a linear relationship -> showing this regression is not a good fit. 

```{r analyse residuals of lin. regression of all tripartite buildings}
# following the hints of https://www.r-bloggers.com/visualising-residuals/

# 1. all tripartite buildings

fit <- lm(summe_aufwand ~ summe_volumen, data = aufwand_MSG)  # Fit the model: predicted value ~ x-Achsen-value

aufwand_MSGfit <- aufwand_MSG

aufwand_MSGfit$predicted <- predict(fit)   # Save the predicted values
aufwand_MSGfit$residuals <- residuals(fit) # Save the residual values

ggplot(aufwand_MSGfit, aes(x = summe_volumen, y = summe_aufwand)) + 
  geom_point() +
  geom_point(aes(x = summe_volumen, y = predicted), shape = 1) +  # Add the predicted values
  geom_segment(aes(xend = summe_volumen, yend = predicted), # lines between predicted and true values
               alpha = .2) +  # alpha to fade lines  
  geom_point(aes(color = residuals)) +  # Color mapped here
  scale_color_gradient2(low = "blue", mid = "white", high = "red") +  # Colors to use here
  guides(color = FALSE) + # we dont need a legend
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey")   # Plot regression slope

summary(fit)

plot(fit)
```

2. for the tripartite buildings without the stone-cone building

First plot visualises the residuals on the original plot. Then typical residual-analysis. 

Conclusion: much better, resdiduals much smaller and not in a linear relationship with fitted values. 

```{r residual analysis of linear regression for tripartite buildings without the stone cone building}


fit2 <- lm(summe_aufwand ~ summe_volumen, data = aufwand_MSG_oSSG)  # Fit the model

aufwand_MSG_oSSGfit <- aufwand_MSG_oSSG

aufwand_MSG_oSSGfit$predicted <- predict(fit2)   # Save the predicted values
aufwand_MSG_oSSGfit$residuals <- residuals(fit2) # Save the residual values


ggplot(aufwand_MSG_oSSGfit, aes(x = summe_volumen, y = summe_aufwand)) +
  geom_point() +
  geom_point(aes(x = summe_volumen, y = predicted), shape = 1) +  # Add the predicted values
  geom_segment(aes(xend = summe_volumen, yend = predicted),
               alpha = .2) +  # alpha to fade lines 
  geom_point(aes(color = residuals)) +  # Color mapped here
  scale_color_gradient2(low = "blue", mid = "white", high = "red") +  # Colors to use here
  guides(color = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey")+           # Plot regression slope
       
  geom_label_repel(aes(x = summe_volumen,  # label the points
                        y = summe_aufwand,
                       label = uebersicht.X),
                   box.padding = 0.35,             # bounding box of label
                   point.padding = 0.5,
                   size = 5,
                   nudge_x = 0,
                   nudge_y = 0.05,                 
                   direction = "y")

summary(fit2)

plot(fit2)

```

3. How much does the Stone-Cone Building differ from the "other" tripartite building linear model?
```

# first: get the two values: the expected value of the stone-cone building on the linear model calculated without it
expSSGaufwand <- -112 + 3.44*aufwand_MSG[9,12] # this is the formula of the linear model without the stone-cone building times the mass to get to expected effort
realSSGaufwand <- aufwand_MSG[9,22] # this is the value of the real effort

# two values for comparison
diffSSGaufwand<- realSSGaufwand - expSSGaufwand
divSSGaufwand <- realSSGaufwand/expSSGaufwand

```

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository("../..")
```
